Parsing Postfix log files

### 2009-03-11 doc/logparser-architecture.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

Rules are responsible for categorising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a regex when recognising inputs, but other approaches may
prove useful for some applications, e.g.\ comparing fixed strings to the
input, or checking the length of the input; for the remainder of this
thesis it will be assumed that a regex is used.  Each rule must specify at
least the regex to recognise inputs and the action to invoke when
recognition is successful, but implementations are free to add any other
attributes they require; \sectionref{rule attributes} describes the
attributes used in \parsername{}, and some generally useful attributes will
be discussed later in this section.

>>>>

Rules are responsible for categorising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a regex when recognising inputs, but other approaches may
prove useful for some applications, e.g.\ comparing fixed strings to the
input, or checking the length of the input; for the remainder of this
thesis it will be assumed that a regex is used.  Each rule must specify at
least the regex to recognise inputs and the action to invoke when
recognition is successful, but implementations are free to add any other
attributes they require; \sectionref{rules table} describes the attributes
used in \parsername{}, and some generally useful attributes will be
discussed later in this section.

### 2009-03-11 doc/logparser-implementation.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

\label{rule attributes}

@@

%\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]

XXX IMPROVE HOW THIS LOOKS

@@

\subsection{Actions in Detail}

@@

XXX ONCE THIS PARAGRAPH HAS BEEN INTEGRATED AND EXPANDED UPDATE THE
REFERENCE IN ARCHITECTURE->RULES->ATTACHING CONDITIONS TO RULES\@.
Each log line contains the name of the Postfix component that produced it,
and this information is used when determining which rules should be used to
parse each log line (see \sectionref{rules in implementation} for details).

The parser does not try to detect overlapping rules; that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the order in which rules
are tried against each line may change between log files, and the first
matching rule wins.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
specific sites from \acronym{SMTP} delivery to the rest of the world.  The
algorithm provides a mechanism for ordering overlapping rules: the priority
field in each rule (defaults to zero).  Negative priorities may be useful
for catchall rules.

    \item Sort by regex and visually inspect the list, e.g.\ with
        \acronym{SQL} similar to: \textbf{select regex from rules order by
        regex;}

    \item Compare the results of parsing using sorted, shuffled, and
        reversed rules (described in \sectionref{rule ordering for
        efficiency}).  Parse several log files using optimal ordering, then
        dump a textual representation of the rules, connections, and
        results tables.  Repeat with shuffled and reversed ordering,
        starting with a fresh database.  If there are no overlapping rules
        the tables from each run will be identical; differences indicate
        overlapping rules.  The rules that overlap can be determined by
        examining the differences in the tables: each result contains a
        reference to the rule which created it, if the references differ
        between runs the two rules referenced in the differing results
        overlap.  Unfortunately this method cannot prove the absence of
        overlapping rules; it can detect overlapping rules, but only if
        there are log lines in the log files that are matched by more than
        one rule.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address because the domain in the sender address does
not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
to the sender's address (for full details
see~\cite{reject-unknown-sender-domain}).

Example log line matched by this rule:

@@

\begin{verbatim}
NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550 5.1.8
  <foo@example.com>: Sender address rejected: Domain not found;
  from=<foo@example.com> to=<info@example.net>
  proto=SMTP helo=<smtp.example.com>
\end{verbatim}

% do not reformat this!
\begin{tabular}[]{ll}

@@

\vspace{1em}

Fields matched by the regex will be automatically saved to the results and
connections tables.

@@

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's regex will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.

    \item [action:] The action to be executed if the regex matches
        successfully.  See \sectionref{actions in detail in implementation}
        for details of the actions implemented by the parser,
        \sectionref{actions in architecture} for the role of actions in the
        parser architecture.

    \item [result\_data and connection\_data:] are used to provide XXX

    \item [hits, hits\_total, and priority:] hits and priority are used in
        ordering the rules (see \sectionref{rule ordering for efficiency});
        hits is set to the number of successful matches at the end of the
        parsing run; hits\_total is the sum of hits over every parsing run,
        but is otherwise unused by the parser.

%    \item [cluster\_group] The cluster\_group attribute is used by the
%        Decision Tree algorithm described in a separate document; the
%        parser does not use it in any way.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different
\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
facilitate parsing new log lines, the parser's design separates parsing
rules from parsing actions: adding new actions can be difficult, but adding
new rules to parse new rejection messages is trivial, and occurs much more
frequently.  A program is provided to ease the process of creating new
rules from unparsed log lines, based on the algorithm developed by Risto
Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
the two algorithms will be outlined as part of the general explanation
below.

The core of the \acronym{SLCT} algorithm is quite simple: log lines are
generally created by substituting variable words into a fixed pattern, and
analysis of the frequency with which each word occurs can be used to
determine whether the word is variable or part of the fixed pattern.  This
classification can be used to group similar log lines and generate a regex
to match each group of log lines.

There are 4 steps in the algorithm:

@@

    \item [Pre-process the file]  The new algorithm leverages the knowledge
        gained when writing rules and performs a large number of
        substitutions on the input log lines, replacing commonly occurring
        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
        standard start of rejection messages, etc.) with regex keywords
        that the parser will expand when it loads the rule (see the
        regex entry in \sectionref{rule attributes}).  The purpose of
        this step is to utilise existing knowledge to create more accurate
        regexes; it replaces a large number of variable words with fixed
        words, improving the subsequent classification of words as fixed or
        variable.  The new log lines are written to a temporary file, which
        all subsequent stages use instead of the original input file.

        In the original algorithm the purpose of the pre-processing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
        small range of values for each word of each log line, incrementing
        a counter for each hash.  The counters will be used in the next
        pass to filter out words: if the word's hash does not have a high
        frequency, the word itself cannot have a high frequency, so there
        is no need to maintain a counter for it, reducing the number of
        counters required and thus the program's memory consumption.

    \item [Calculate word frequencies]  The position of words within a log
        line is important: a word occurring in two log lines does not
        indicate similarity unless it occupies the same position within
        both log lines.\footnote{If a variable term within a line contains
        spaces, it will appear to the algorithm as two or more words rather
        than one.  This will alter the position of subsequent words, so a
        word occurring in different positions in two log lines
        \textit{may\/} indicate similarity, but the algorithm does not
        attempt to deal with this possibility.}  The algorithm maintains a
        counter for each \texttt{(word, word's position within the log
        line)} tuple, incrementing it each time that word occurs in that
        position.

        The original algorithm only maintains counters for words whose hash
        result from the previous step has a high frequency; this reduces
        the number of counters maintained by the algorithm, reducing the
        memory requirements of the algorithm at a cost of increased CPU
        usage.

        As time goes on the amount of memory typically available increases
        and the requirement to reduce memory requirements decreases, hence
        the modified algorithm omits the hashing step and maintains
        counters for all tokens.  In addition most of the infrequently
        occurring words will have been substituted with keywords during the
        first step, vastly reducing the number of tuples to maintain
        counters for.

    \item [Classify words based on their frequency]  The frequency of every
        tuple \texttt{(word, word's position within the log line)} is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point)
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new regexes.  Variable terms are
        replaced by \texttt{.+}, which means to match zero or more of any
        character.

    \item [Build regexes]  The words are reassembled to produce a regex
        matching the log line, and a counter is maintained for each
        regex.  Contiguous sequences of \texttt{.+} in the newly
        reassembled regexes are collapsed to a single \texttt{.+}; any
        resulting duplicate regexes are combined, and their counters
        added together.  If a regex's counter is lower than the
        threshold supplied by the user the regex is discarded; this
        second threshold is independent of the threshold used to
        differentiate between fixed and variable words, but once again 1\%
        of log lines is a good starting point.  The new regexes are
        displayed for the user to add to the database, either as new rules
        or merged into the regexes of existing rules; the counter for
        each regex is also displayed, giving the user an indication of
        how many of the log lines that regex should match.  Discarding
        regexes will result in some of the log lines not being matched;
        when the rules have been augmented with the new regexes, the
        original log files should be parsed again, and any remaining
        unparsed log lines used as input to this utility.

@@

A second utility is also provided that reads a list of new regexes and
the input given to the first utility.  It tries to match each input log
line against each regex, counting the number of log lines that match
each regex, warning the user if an input log line is matched by more
than one regex, and additionally warning if an input log line is not
matched by any regex.  It displays a summary of how many input log lines
each regex matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the regexes
produced by the first utility are correctly matching the input log lines
they are based upon.  A future version of this utility will also group
input log lines by regex, so the user can tweak the regexes if
required.

@@

>>>>

\label{rules table}

@@

@@

\subsection{Actions in the \parsernamelong{}}

@@

The rules are responsible for classifying each log line and specifying the
correct action to be invoked.  For any parser implemented using this
architecture, the rules are the most visible component and most likely to
be modified by users.  Rules need to be as simple as possible so that users
can easily modify or add to them, but the implementation must balance that
simplicity with the need to provide enough flexibility and power to
successfully parse the inputs.

The theoretical aspects of this component are covered in detail in
\sectionref{rules in architecture}; this section is concerned with the
practical aspects of how rules are implemented and used in \parsername{}.
The structure of the rules table has already been documented in
\sectionref{rules table}; that abstract description will not be repeated
here, but should be fleshed out by the concrete example rule in
\sectionref{example rule in implementation}.  The process of creating new
rules from unparsed log lines is dealt with in \sectionref{creating new
rules in implementation}; the implementation of the utility supplied with
\parsername{} to create regexes from unparsed log lines is also described
in that section.  The final portion of this section provides some
suggestions for how to detect overlapping rules.

@@

The example rule in \tableref{Example rule in implementation table}
recognises the log line resulting from Postfix rejecting a delivery attempt
because the domain in the sender address does not have an A, AAAA,
or MX DNS entry, i.e.\ mail could not be delivered to the sender's address
(for full details see~\cite{reject-unknown-sender-domain}).

Example log line recognised by this rule:

@@

\begin{verbatim}
NOQUEUE: reject: RCPT from smtp.example.com[192.0.2.1]:
  550 5.1.8 <foo@example.com>:
  Sender address rejected: Domain not found;
  from=<foo@example.com> to=<info@example.net>
  proto=SMTP helo=<smtp.example.com>
\end{verbatim}

% do not reformat this!
\begin{table}[ht]

    \caption{Example rule}
    \empty{}\label{Example rule in implementation table}
    \begin{tabular}{ll}

@@

\end{table}

The fields in \tableref{Example rule in implementation table} are used as
follows:

@@

    \item [program and regex:] If the program in the rule equals the
        program in the log line, a match using the rule's regex will be
        attempted against the log line; if the match is successful the
        action will be executed, if not the next rule will be tried.  If
        the program-specific rules do not match the log line, the parser
        will fall back to generic rules; if those rules are unsuccessful a
        warning will be issued.  Data captured by the regex (sender,
        recipient, etc.) will be automatically saved to the results and
        connections tables.

    \item [action:] The action to be invoked when the rule recognises a log
        line.  See \sectionref{actions in detail in implementation} for
        details of the actions implemented by \parsername{}, and
        \sectionref{actions in architecture} for the role of actions in the
        parser architecture.

    \item [result\_data and connection\_data:] are used to provide data not
        present in the log line, and are unused in this rule.

    \item [hits, hits\_total, and priority:] hits and priority are used
        when ordering the rules for more efficient parsing (see
        \sectionref{rule ordering for efficiency}).  At the end of each
        parsing run hits is set to the number of log lines recognised by
        the rule.  Hits\_total is the sum of hits over every parsing run,
        but is otherwise unused by the parser.

@@

The log files produced by Postfix differ from installation to installation,
because administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different
\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
facilitate parsing new log lines, the architecture separates rules from
actions: adding new actions can be difficult, but adding new rules to
recognise new log lines is trivial, and occurs much more frequently.

XXX DESCRIBE HOW TO GO ABOUT ADDING A NEW RULE\@.

A program is provided with \parsername{} to ease the process of creating
new rules from unrecognised log lines, based on the algorithm developed by
Risto Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences
between the two algorithms will be outlined as part of the general
explanation below.  The core of the \acronym{SLCT} algorithm is quite
simple: log lines are generally created by substituting variable words into
a fixed pattern, and analysis of the frequency with which each word occurs
can be used to determine whether the word is variable or part of the fixed
pattern.  This classification can be used to group similar log lines and
generate a regex to match each group of log lines.  There are 4 steps in
the algorithm:

@@

    \item [Pre-process the file.]  The new algorithm leverages the
        knowledge gained when writing rules and developing \parsername{};
        it performs a large number of substitutions on the input log lines,
        replacing commonly occurring variable terms (e.g.\ email addresses,
        \acronym{IP} addresses, the standard start of rejection messages)
        with keywords that \parsername{} will expand when it loads the new
        rule.  The purpose of this step is to utilise existing knowledge to
        create more accurate regexes; it replaces a large number of
        variable words with fixed words, improving the subsequent
        classification of words as fixed or variable.  The altered log
        lines are written to a temporary file, which subsequent stages will
        use instead of the original input file.

        In the original algorithm the purpose of the pre-processing stage
        was to reduce the memory consumption of the program.  In the first
        pass it generates a hash~\cite{hash-functions}\glsadd{hash} (from a
        small range of values) for each word of each log line, incrementing
        a counter for each hash.  The counters will be used in the next
        pass to filter out words: if the word's hash does not have a high
        frequency, the word itself cannot have a high frequency, so there
        is no need to maintain a counter for it, reducing the number of
        counters required and thus the program's memory consumption.

    \item [Calculate word frequencies.]  The position of words within a log
        line is important: a word occurring in two log lines does not
        indicate similarity unless it occupies the same position in both
        log lines.  If a variable term within a line contains
        spaces, it will appear to the algorithm as more than one word.
        This will alter the position of subsequent words, so a word
        occurring in different positions in two log lines \textit{may\/}
        indicate similarity, but the algorithm does not attempt to deal
        with this possibility.  The algorithm maintains a counter for each
        \texttt{(word, word's position within the log line)} tuple,
        incrementing it each time that word occurs in that position.

        The original algorithm hashes each word and checks that result's
        counter from the previous pass; if that counter has a high
        frequency a separate counter will be maintained for this word in
        this position.  This reduces the number of counters maintained by
        the algorithm in this step, reducing the memory requirements of the
        algorithm at a cost of increased CPU usage.

        As time goes on the amount of memory typically available increases
        and the need to reduce memory requirements decreases, so the
        modified algorithm omits the hashing step and maintains counters
        for all tokens.  In addition most of the infrequently occurring
        words will have been substituted with keywords during the first
        step, vastly reducing the number of tuples to maintain counters
        for.

    \item [Classify words based on their frequency.]  The frequency of
        every tuple \texttt{(word, word's position within the log line)} is
        checked: if its frequency is greater than the threshold supplied by
        the user (1\% of all log lines is generally a good starting point),
        it is classified as a fixed word, otherwise it is classified as a
        variable term.  If a variable term appears sufficiently often it
        will be misclassified as a fixed term, but that should be noticed
        by the user when reviewing the new regexes.  Variable terms are
        replaced by \texttt{.+} to match one or more of any character.

    \item [Build regexes.]  The words are reassembled to produce a regex
        matching the log line, and a counter is maintained for each regex.
        Contiguous sequences of \texttt{.+} in the newly reassembled
        regexes are collapsed to a single \texttt{.+}; any resulting
        duplicate regexes are combined, and their counters added together.
        If a regex's counter is lower than the threshold supplied by the
        user the regex is discarded; this second threshold is independent
        of the threshold used to differentiate between fixed and variable
        words, but once again 1\% of log lines is a good starting point.
        The new regexes are displayed for the user to add to the database,
        either as new rules or merged into the regexes of existing rules;
        the counter for each regex is also displayed, giving the user an
        indication of how many of the log lines that regex should match.
        Discarding regexes will result in some of the log lines not being
        matched; when the rules have been augmented with the new regexes,
        the original log files should be parsed again, and any remaining
        unparsed log lines used as input to this utility.

@@

A second utility is also provided that reads a list of new regexes and
the input given to the first utility.  It tries to match each input log
line against each regex, counting the number of log lines that match
each regex, warning the user if an input log line is matched by more
than one regex, and additionally warning if an input log line is not
matched by any regex.  It displays a summary of how many input log lines
each regex matched, comparing it to the expected number of matches; this
provides the user with an easy method of checking if the regexes
produced by the first utility are correctly matching the input log lines
they are based upon.

@@

\parsername{} does not try to detect overlapping rules; that responsibility
is left to the author of the rules.  The advantages and difficulties of
overlapping rules have already been addressed in \sectionref{overlapping
rules in architecture} and will not be repeated here.  \parsername{}
provides a mechanism for ordering overlapping rules: the priority field in
each rule (defaulting to zero); negative priorities may be useful for
catchall rules.

    \item Sort by program and regex, then visually inspect the list, e.g.\
        with \acronym{SQL} similar to:      \newline{}
        \verb!SELECT program, regex!        \newline{}
        \verb!    FROM rules!               \newline{}
        \verb!    ORDER BY program, regex;! \newline{}
        The rules are sorted first by program, then by regex, because rules
        with overlapping regexes will not overlap if their programs are
        different.  Note that this does not properly deal with rules whose
        program is \texttt{*}; those rules will be used on all log lines
        that have not been recognised by program-specific rules.

    \item Compare the results of parsing using different rule orderings
        (described in \sectionref{rule ordering for efficiency}).  Parse
        several log files using optimal ordering, then dump a textual
        representation of the rules, connections, and results tables.
        Repeat with shuffled and reversed ordering, starting with a fresh
        database.  If there are no overlapping rules the tables from each
        run will be identical; differences indicate overlapping rules.  The
        rules that overlap can be determined by examining the differences
        in the tables: each result contains a reference to the rule which
        created it, which will change if that rule overlaps with another.
        Unfortunately this method cannot prove the absence of overlapping
        rules; it can detect overlapping rules, but only if there are log
        lines in the log files that are recognised by more than one rule.

### 2009-03-11 doc/logparser-other-parsers-review.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

    \item The parsers reviewed all produce a report of varying complexity
        and detail, whereas \acronym{PLP} does not; it extracts data and
        leaves generation of reports from the data to other programs.
        Using an \acronym{SQL} database simplifies the process of
        generating such reports (discussed in \sectionref{database as
        API}); some sample queries are given in \sectionref{motivation}.
        The parser developed for this project is designed to enable much
        more detailed log file analysis by providing a stable platform for
        subsequent programs to develop upon.

>>>>

    \item The parsers reviewed all produce a report of varying complexity
        and detail, whereas \parsername{} does not; it extracts data and
        leaves generation of reports from the data to other programs.
        Using an \acronym{SQL} database simplifies the process of
        generating such reports (discussed in \sectionref{database as
        API}); some sample queries are given in \sectionref{motivation}.
        The parser developed for this project is designed to enable much
        more detailed log file analysis by providing a stable platform for
        subsequent programs to develop upon.
